Assessing primitives performance\\ on multi-stage execution

### 2018-05-16 paper.tex
replace interProxy by interface, take some other comments into account

\cba{We need to introduce here the concept of VM plugin and interpreterproxy indirection/overhead. This is very important since later we just say this is a plugin and we don't redefine it}
\sk{done but someone better check what I wrote :-)}

Cog VM can be extended using plugins \cite{Guzd01b}.
VM plugins enable the addition of new features to the VM. There are two types of plugins:  the external and the  built-ins ones.
\begin{itemize}
\item \emph{External plugins:} Plugins that can be added or removed from the VM without the need to recompile the VM sources.
\item \emph{Internal plugins:} Plugins that cannot be installed without recompiling the VM sources. However, these plugins can reach some parts of the VM internals\footnote{as data structures, routines...} that remains unreachable from an external plugin. These operations are performed via an InterpreterProxy. Although, this solution results in a substential overhead if called often. \cba{Remove from paper InterpreterProxy}
\end{itemize}

@@

In addition to the previous implementation, we've re-implemented here the primitive on top of Cog's JIT back-end. The primitive is written in Cog's RTL, an abstract assembly whose instructions compile almost one-to-one to native instructions. This primitive requires a more important amount of work to be written: the number of line of code is larger, the implementor has to be careful to generate efficient machine code (there is no C compiler to optimize the code) and debugging such code requires the use of a processor simulator \sd{such as Boch}. Since only performance critical implementations matter in this context, we've only re-implemented the case where there are two parameters. For the three parameter primitive, the VM falls back to the Slang version. This primitive follows the new specification, the full code is available in Appendix \ref{RTLAppendix}.

@@

 In this section we compare the five different implementations according to the following criteria:
 \begin{itemize}
 	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes). \sd{why not ST, Plugin, VM?}
	\item \emph{Implementation language familiar to developers:} Smalltalk application developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code. Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.  We use All if all communities are familiar with the language, VM if all VM developers are and VM-JIT if only a subset of VM developers are.
	\item \emph{Readability:} To evaluate the approach readability, we use the number of lines of code even if this is arguable but used regularly in other papers \cite{xxx}. A better metric would be to use the cyclomatic complexity, however since the  Cog' s RTL generates branches at runtime, it is difficult to use it.
	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to be executed if the primitive fails. This code is maintained by the client language implementors, and not by the VM implementors. Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation\sd{why?}. However, as a VM implementor, we need to maintain what generates the C code of the VM, hence, in the case of Smart Syntax, we need to maintain a different method in each client language. We also count all the Slang implementations as well as the ones on Cog's RTL.
	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
	\item \emph{Optimization potential:} \cba{need to mention that this is for large strings}
	C cross-file inlining is not performed by default for internal plugins. External plugins don't have access to the whole VM API and have to go through InterpreterProxy to do so, leading to an overhead. \cba{InterpreterProxy not defined before}
	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code. The generated code can therefore differ according the compilers and the optimizations they performed. On the contrary, the JIT will always generate the same native code.
\end{itemize}

@@

In each benchmark, we run a string comparison in a loop and measure the total execution time.
The number of iterations of the loop depends on the string size: the comparison is repeated between 1414 and 10000 times for a short string (5 characters or less), and between 100 and 316 times for a long string (100 characters or more). This addresses the warm-up of the JIT \cba{??? This needs to be explained differently. We get into jitted code after a small number of loop iteration (I think 20 currently) or second call. The JIT overhead is negligible here for 1 second because we could not see any time spent in the JIT in the profiler + ref to your paper, i.e. less than 0.5ms in the JIT} \cite{CogProfiler} and allows the benchmarks to run at least 1 second in the slower implementations \cba{Ok, 1 second to avoid processor / OS noises, not just for fun}.
The loop overhead is measured and removed from the measurements.

@@

\paragraph{Stack switch overhead.}
As mentioned in Section~\ref{sec:VMexec}, all implementations written in Slang come with an overhead due to the switch between the C and Smalltalk stacks. This overhead is clearly visible in the bench results in Figure~\ref{fig:bench} for small strings (size 1 or less): the SmartSyntax and Plugin implementations are around 30 to 60\% slower than the baseline's, and the PureSlang performance is close to the baseline's.
As the string size goes larger, the overhead is less significant because it is absorbed by other performance gains.
Regarding this criteria, the Slang+JIT version is always the fastest: it is not really impacted by the switch overhead, because the Slang part is used for a very short time \sd{Why that? just during the time the JIT does not kick in?}.

>>>>

Cog VM can be extended using plugins \cite{Guzd01b}.
VM plugins enable the addition of new features to the VM. There are two types of plugins:  the external and the internal ones.
\begin{itemize}
\item \emph{External plugins:} Plugins that can be added or removed from the VM without the need to recompile the VM sources.
\item \emph{Internal plugins:} Plugins that cannot be installed without recompiling the VM sources. However, these plugins can reach via an interface some parts of the VM internals such as data structures, routines,... that remains unreachable from an external plugin. This solution results in a substential overhead if called often. 
\end{itemize}

@@

In addition to the previous implementation, we've re-implemented here the primitive on top of Cog's JIT back-end. The primitive is written in Cog's RTL, an abstract assembly whose instructions compile almost one-to-one to native instructions. This primitive requires a more important amount of work to be written: the number of line of code is larger, the implementor has to be careful to generate efficient machine code (there is no C compiler to optimize the code) and debugging such code requires the use of a processor simulator such as Bochs. Since only performance critical implementations matter in this context, we've only re-implemented the case where there are two parameters. For the three parameter primitive, the VM falls back to the Slang version. This primitive follows the new specification, the full code is available in Appendix \ref{RTLAppendix}.

@@

 In this section we compare the five different implementations according to the following criteria:
 \begin{itemize}
 	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes). \sd{why not ST, Plugin, VM?}
	\item \emph{Implementation language familiar to developers:} Smalltalk application developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code. Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.  We use All if all communities are familiar with the language, VM if all VM developers are and VM-JIT if only a subset of VM developers are.
	\item \emph{Readability:} To evaluate the approach readability, we use the number of lines of code even if this is arguable but used regularly in other papers \cite{xxx}. A better metric would be to use the cyclomatic complexity, however since the  Cog' s RTL generates branches at runtime, it is difficult to use it.
	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to be executed if the primitive fails. This code is maintained by the client language implementors, and not by the VM implementors. Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation\sd{why?}. However, as a VM implementor, we need to maintain what generates the C code of the VM, hence, in the case of Smart Syntax, we need to maintain a different method in each client language. We also count all the Slang implementations as well as the ones on Cog's RTL.
	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
	\item \emph{Optimization potential on large string comparison:}
	C cross-file inlining is not performed by default for internal plugins. External plugins don't have access to the whole VM API and have to go through an interface to do so, leading to an overhead.
	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code. The generated code can therefore differ according the compilers and the optimizations they performed. On the contrary, the JIT will always generate the same native code.
\end{itemize}

@@

In each benchmark, we run a string comparison in a loop and measure the total execution time.
The number of iterations of the loop depends on the string size: the comparison is repeated between 1414 and 10000 times for a short string (5 characters or less), and between 100 and 316 times for a long string (100 characters or more). This addresses the warm-up of the JIT \cba{??? This needs to be explained differently. We get into jitted code after a small number of loop iteration (I think 20 currently) or second call. The JIT overhead is negligible here for 1 second because we could not see any time spent in the JIT in the profiler + ref to your paper, i.e. less than 0.5ms in the JIT} \cite{CogProfiler} and allows the benchmarks to run at least 1 second in the slower implementations to avoid processor / OS noises.
The loop overhead is measured and removed from the measurements.

@@

\paragraph{Stack switch overhead.}
As mentioned in Section~\ref{sec:VMexec}, all implementations written in Slang come with an overhead due to the switch between the C and Smalltalk stacks. This overhead is clearly visible in the bench results in Figure~\ref{fig:bench} for small strings (size 1 or less): the SmartSyntax and Plugin implementations are around 30 to 60\% slower than the baseline's, and the PureSlang performance is close to the baseline's.
As the string size goes larger, the overhead is less significant because it is absorbed by other performance gains.
Regarding this criteria, the Slang+JIT version is always the fastest as it is not impacted by the switch overhead.
